---
title: "Causal Analysis: Study Load on Anxiety"
output: html_notebook
---

Constructing the DAG, the following were decided:

Outcome: Anxiety Levels(0 to 21)
Treatment: Having High Study Load
Treatment Effect: Effect of Study Load on Anxiety Levels
Treatment Group: Students having high Study Load (Study Load = 3,4,5)
Control Group: Students having lower study load (Study Load = 0,1,2)

```{r}
## Installing the Required Packages
library('dplyr')
library('magrittr')
library('randomizr')
library('labelled')
library('marginaleffects')
library("MatchIt")
library("ggplot2")
library("tidyverse")

```

Using the "Student Stress Factors" dataset, we attempt to find if study load has an impact on a student's anxiety levels.


```{r}
data = read.csv("StressLevelDataset.csv")
head(data)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

```{r}
summary(data)
```

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

```{r}
install.packages("dagitty")
library(dagitty)

# Define your DAG
dag <- dagitty('dag {
  "High Study Load" -> "Anxiety Levels"
  "Self Esteem" -> "Anxiety Levels"
  "Social Support" -> "Anxiety Levels"
  "Sleep Quality" -> "Anxiety Levels"
  "Physical Activity" -> "Anxiety Levels"
  "High Study Load" -> "Self Esteem"
  "Social Support" -> "Sleep Quality"
}')

# Plot the DAG
plot(dag)
```
